---
title: "Exploring item responses"
date: "`r Sys.Date()`"
author: Kristine Zheng
output:
  html_document:
    toc: true
    toc_float: true
    theme: paper
    code_folding: "hide"
    df_print: paged
---

## Set up
```{r}
rm(list=ls())
```

```{r setup, include=FALSE, warning=F, message=F}
if(!require("pacman")) install.packages("pacman")
pacman::p_load('tidyverse','here','tidyverse', 'mosaic')
here::i_am("analysis/02_item_responses.Rmd")
options(digits=5)
set.seed(12341)
knitr::opts_chunk$set(
	fig.width = 6,
	fig.height = 4,
	message = FALSE,
	warning = FALSE
)
```

## Load data

```{r}
INPUT_PATH = file.path( '..', 'data', 'college-23') # top-level directory for data
OUTPUT_PATH = file.path( '..', 'data', 'college-23-processed') # top-level directory for data
RESPONSES_PATH = file.path(INPUT_PATH, 'raw', 'responses.Rdata')
CLASSES_PATH = file.path(INPUT_PATH, 'classes.csv')

# TODO: figure out which codebooks are relevant
CODEBOOK_PAGEVIEWS_PATH = file.path(INPUT_PATH, 'codebooks', 'codebook_page_pageviews.csv') 
CODEBOOK_PAGE_PATH = file.path(INPUT_PATH, 'codebooks', 'codebook_page.csv')
CODEBOOK_SURVEYS_PATH = file.path( '..', 'data', 'codebooks', 'survey-codebooks', 'codebook_all_surveys.csv')
```

```{r}
# Load raw data from college-23 repository
# NB: takes 20-30s
load(file.path(RESPONSES_PATH)) # loads `responses` dataframe
glimpse(responses)
```

```{r}
codebook_classes = read_csv(file.path(CLASSES_PATH)) # matches institutions and classes with textbooks and number of students 
codebook_pageviews = read_csv(file.path(CODEBOOK_PAGEVIEWS_PATH)) |> select(-1) # codebook for matching chapter and page information, removes non-textbook (survey) pages
codebook_page = read_csv(file.path(CODEBOOK_PAGE_PATH)) |> select(-1) # details the types and count of items on each page (also a chapter version)

codebook_all_surveys <- read.csv(file.path(CODEBOOK_SURVEYS_PATH))

glimpse(codebook_pageviews)
```

## Plan
Overarching goal: gather response level summary analyses and characterize variation between students, classes, chapters, etc 

step 1) aggregate information about attempts, time, etc from the response level to the item level
- filter out the survey items (merge with codebook_pageviews)
step 2) distributions
- items: sorted by easy to hard (proportion correct when attempted) and colored by chapter
- student level (colored by class/chapter): attempts, proportion completed/attempted, proportion correct, proportion correct when completed, proportion correct when attempted 
- chapter level: instances of an item
- class level: proportion completed, chapters accessed 

df:
- items (separated by student) --> aggregated overall across all students 

## STEP 1: PROCESS DATA
### 1) Filter responses 
```{r}
filtered_responses = responses |>
  left_join(
    codebook_pageviews,
    by = c('release', 'book', 'chapter', 'page')
  ) |>
  filter(!is.na(page_num)) |> # remove survey pages and other pages 
  filter(!(item_id %in% codebook_all_surveys$item_id)) |> 
  filter(!str_detect(page, "Review Questions")) # eoc quiz questions
  #filter(str_detect(page, "Review Questions")) # eoc quiz questions
```

```{r}
# check which one is unique item identitfier: lrn_activity_reference, lrn_question_reference, item_id
n_distinct(filtered_responses$item_id) # 840 --> not uniaue but use this because ck uses this
n_distinct(filtered_responses$lrn_activity_reference) # 843 --> definitely not unique for each question
n_distinct(filtered_responses$lrn_question_reference) # 910 --> 
n_distinct(filtered_responses$prompt) # 1172 --> not unique 
```

TODO: investigate why item_id doesn't match number of questions?
```{r}
# check which one is unique item identitfier: lrn_activity_reference, lrn_question_reference, item_id
filtered_responses |>
  distinct(item_id, prompt, .keep_all = TRUE) |>
  select(item_id, prompt) |>
  nrow() # 1250

filtered_responses |>
  distinct(item_id, lrn_activity_reference, lrn_question_reference, .keep_all = TRUE) |>
  select(release, chapter_num, item_id, lrn_activity_reference, lrn_question_reference, prompt) |>
  View()

filtered_responses |>
filter(item_id == "Ch2_Starting_1") |>
  distinct(student_id, item_id, lrn_activity_reference, lrn_question_reference, prompt) |>
  View()
```

### 2) Aggregate response attempts to items 
-> filter out actual content (TODO KRISTINE TO CHECK ON THIS)
-> use first correct attempt 

```{r}
items = filtered_responses |>
  group_by(class_id, student_id, chapter_num, page_num, item_id) |> # group by unique items per student 
  arrange(attempt) |>
  drop_na(points_possible, points_earned) |>
  mutate(
     correct_first_try = ifelse(attempt == 1, points_earned, NA),
     is_correct_ever = max(points_earned, na.rm = TRUE), # why is this sometimes Na
     n_attempts = max(attempt, na.rm = TRUE),
     dt_submitted_list = list(dt_submitted),
     first_correct_attempt = ifelse(any(points_earned == points_possible, na.rm = TRUE),  
                                   min(attempt[points_earned == points_possible], na.rm = TRUE),  
                                   NA_real_),  # Get first attempt where correct, else NA
     correctness_over_attempts = is_correct_ever / first_correct_attempt )|>
  ungroup() |>
  select(book, release, class_id, student_id, chapter_num, page_num, item_id, item_type,
         correct_first_try, is_correct_ever, points_possible, first_correct_attempt, n_attempts, correctness_over_attempts, dt_submitted_list, lrn_status) |>
  distinct() 

#View(items)
```

save data 
```{r}
# Save as .RData
save(items, file = file.path(OUTPUT_PATH, 'processed_college_23_items.RData'))
# Save as csv
write_csv(items, file.path(OUTPUT_PATH, 'processed_college_23_items.csv'))

```

## STEP 2: VISUALIZE

### 0) class variation
-> chapters accessed per class
```{r}
n_distinct(items$class_id) # 48 
```
chapters accessed per class
```{r}

class_chapters <- items |>
  group_by(class_id) |>
  arrange(class_id, chapter_num) |>
  summarise(first_chapter = first(chapter_num), 
            last_chapter = last(chapter_num),
            chapters_accessed = n_distinct(chapter_num)
            ) |> 
  mutate(
    class_id = factor(class_id, levels = class_chapters$class_id[order(class_chapters$last_chapter)])
    )

# Create the plot with lines indicating first and last chapters accessed by each class
ggplot(class_chapters) +
  geom_segment(aes(x = first_chapter, xend = last_chapter, y = class_id, yend = class_id), 
               color = "gray", size = 1) +  # Line connecting first and last chapter
  geom_point(aes(x = first_chapter, y = class_id), color = "blue", size = 2) +  # First chapter
  geom_point(aes(x = last_chapter, y = class_id), color = "red", size = 2) +   # Last chapter
  labs(title = "Textbook Access", 
       x = "Chapter", 
       y = "Class") +
  theme_minimal() +
  theme(axis.text.y = element_blank()) +
  scale_x_continuous(breaks = seq(min(class_chapters$first_chapter), 
                                  max(class_chapters$last_chapter), 
                                  by = 1))

favstats(class_chapters$chapters_accessed)
```


```{r}
class_item_completion = items |>
  group_by(class_id) |>
  mutate(
    num_total_attempted = n_distinct(item_id), # multiple item_ids bc everyone does the exercises
    
    total_correct = sum(is_correct_ever, na.rm = TRUE), # number of correct items in a chapter
    num_items = sum(points_possible, na.rm = TRUE), # number of attempted items 
    prop_correct = total_correct/num_items
    )|> 
  ungroup() |> 
  mutate(class_id = factor(class_id, levels = class_chapters$class_id[order(class_chapters$last_chapter)] ))

ggplot(class_item_completion, aes(y = class_id)) +  # Reverse to match ordering visually
  geom_segment(aes(x = 0, xend = num_total_attempted, y = class_id, yend = class_id), 
               color = "gray", size = 1) +  # Line representing range of exercises attempted
  geom_point(aes(x = num_total_attempted, y = class_id), color = "red", size = 2) +   # Last chapter
  labs(title = "", 
       x = "Items Attempted", 
       y = "Class") +
  theme_minimal() +
  theme( axis.text.y = element_blank()) 

# glimpse(class_item_completion)
# hist(class_item_completion$num_total_attempted)

favstats(class_item_completion$num_total_attempted)
```


### 1) items
how many items per chapter?
how many items per student in each chapter 
how many items did each class do?

### 2) completion/attempts (num responses per item)
distribution of attempts (generally, per chapter)
```{r}
# Create the histogram 
ggplot(items, aes(x = n_attempts)) + 
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "", x = "attempts", y = "frequency") +
  theme_minimal()
```

```{r}
chapter_attempt_summary = items |>
  group_by(class_id, student_id, chapter_num) |>
  summarize(
    total_attempts = sum(n_attempts),
    num_items = n(),
    avg_attempt_per_item = total_attempts/ num_items
  )

# plot distribution of attempts per chapter (scatter, box plot)
# Scatter plot: distribution of attempts per chapter
ggplot(chapter_attempt_summary, aes(x = chapter_num, y = avg_attempt_per_item, color = factor(chapter_num))) +
  geom_point(alpha = 0.3) +
  labs(title = "", 
      x = "Chapter", 
       y = "Avg. attempts per item") +
  theme_minimal() +
  theme(legend.position = "none")

# Box plot: distribution of avg. attempts per chapter
ggplot(chapter_attempt_summary, aes(x = factor(chapter_num), y = avg_attempt_per_item)) +
  geom_boxplot(fill = "lightblue", color = "darkblue", alpha = 0.1) +
  labs(title = "", 
       x = "Chapter", 
       y = "Avg. attempts per item") +
  theme_minimal()
```
sort items by avg. number of attempts, color code by chapter 
```{r}
chapter_item_attempt_summary = items |>
  group_by(chapter_num, page_num, item_id) |> # average over students 
  summarize(
    total_attempts = sum(n_attempts, na.rm = TRUE),
    avg_attempts = mean(n_attempts, na.rm = TRUE)
  ) 

glimpse(chapter_item_attempt_summary)

# Plot each item as a line, sorted by average attempts and color-coded by chapter
plot = ggplot(chapter_item_attempt_summary, aes(x = reorder(item_id,total_attempts), 
                         y = total_attempts, 
                         group = item_id, 
                         color = factor(chapter_num))) +
  geom_segment(aes(xend = item_id, yend = 0), size = 1) +  # Vertical line from 0 to avg_attempts
  labs(title = "",
       x = "Item",
       y = "total Attempts Per Item",
       color = "Chapter") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

plot 

# Plot each item as a line, sorted by average attempts and color-coded by chapter
plot = ggplot(chapter_item_attempt_summary, aes(x = reorder(item_id,avg_attempts), 
                         y = avg_attempts, 
                         group = item_id, 
                         color = factor(chapter_num))) +
  geom_segment(aes(xend = item_id, yend = 0), size = 1) +  # Vertical line from 0 to avg_attempts
  labs(title = "",
       x = "Item",
       y = "Average Attempts Per Item",
       color = "Chapter") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

plot 
```

- Proportion completed in whole universe (max number for each class)
- Within class - proportion completed & correct (proportion correct when attempted)


### 3) performance (correctness)
sort items of easy to hard, colored by chapter
variation by class 
within class 
proportion correct out of attempted per class

sort items by avg. number of attempts, color code by chapter 
```{r}
correctness_item_summary = items |>
  group_by(chapter_num, page_num, item_id) |> # average over students 
  summarize(
    total_points = sum(is_correct_ever, na.rm = TRUE),
    avg_points = mean(is_correct_ever, na.rm = TRUE)
  ) 

glimpse(correctness_item_summary)

# Plot each item as a line, sorted by average attempts and color-coded by chapter
plot = ggplot(correctness_item_summary, aes(x = reorder(item_id,avg_points), 
                         y = avg_points, 
                         group = item_id, 
                         color = factor(chapter_num))) +
  geom_segment(aes(xend = item_id, yend = 0), size = 1) +  # Vertical line from 0 to avg_attempts
  labs(title = "",
       x = "Item",
       y = "Average Attempts Per Item",
       color = "Chapter") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

plot 
```


```{r}
chapter_correct_summary = items |>
  group_by(class_id, student_id, chapter_num) |>
  summarize(
    total_correct = sum(is_correct_ever, na.rm = TRUE), # number of correct items in a chapter
    num_items = n(), # number of attempted items 
    prop_correct = total_correct/num_items, # proportion correct per chapter 
  ) |>
  ungroup() 
  
# plot distribution of attempts per chapter (scatter, box plot)
# Scatter plot: distribution of attempts per chapter
ggplot(chapter_correct_summary, aes(x = chapter_num, y = prop_correct, color = factor(chapter_num))) +
  geom_point(alpha = 0.3) +
  labs(title = "", 
      x = "Chapter", 
       y = "prop. correct items") +
  theme_minimal() +
  theme(legend.position = "none")

# Box plot: distribution of avg. attempts per chapter
ggplot(chapter_correct_summary, aes(x = factor(chapter_num), y = prop_correct)) +
  geom_boxplot(fill = "lightblue", color = "darkblue", alpha = 0.1) +
  labs(title = "", 
       x = "Chapter", 
       y = "prop. correct items") +
  theme_minimal()
```

sort items by proportion correct, color code by chapter 
```{r}
chapter_item_correct_summary = items |>
  group_by(chapter_num, page_num, item_id) |> # average over students 
  summarize(
    total_correct = sum(is_correct_ever, na.rm = TRUE), # number of correct items in a chapter
    num_items = n(), # number of attempted items 
    prop_correct = total_correct/num_items, # proportion correct per chapter 
  ) |>
  arrange(prop_correct)

glimpse(chapter_item_correct_summary)

# Plot each item as a line, sorted by prop correct and color-coded by chapter
plot = ggplot(chapter_item_correct_summary, aes(x = reorder(item_id, prop_correct), 
                         y = prop_correct, 
                         color = factor(chapter_num))) +
  geom_segment(aes(x = reorder(item_id, prop_correct), xend = reorder(item_id, prop_correct), y = 0, yend = prop_correct), size = 1) +  
  labs(title = "",
       x = "Item",
       y = "Proportion Correct",
       color = "Chapter") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

plot 
```
#### scatterplot prop correct v avg attempts
```{r}
joined_chapter_summary = chapter_item_attempt_summary |>
  left_join(chapter_item_correct_summary, join_by("chapter_num","page_num","item_id" ))

plot = ggplot(joined_chapter_summary, aes(x = avg_attempts, 
                         y = prop_correct, 
                         color = factor(chapter_num))) +
  geom_point()+
    labs(title = "",
       x = "Avg. Attempts",
       y = "Proportion Students Correct",
       color = "Chapter") +
  theme_minimal() 

plot
cor(joined_chapter_summary$avg_attempts, joined_chapter_summary$prop_correct, use = "complete.obs")

```

```{r}
combined_student_chapter_summary = chapter_correct_summary |>
  left_join(chapter_attempt_summary, join_by("class_id","student_id", "chapter_num", "num_items"))

avg_plot = ggplot(combined_student_chapter_summary, aes(x = avg_attempt_per_item,
                         y = prop_correct,
                         color = factor(chapter_num))) +
  geom_point() +
    labs(title = "",
       x = "Avg. Attempts per item",
       y = "Proportion items Correct",
       color = "Chapter") +
  theme_minimal()

avg_plot
cor(combined_student_chapter_summary$avg_attempt_per_item, combined_student_chapter_summary$prop_correct, use = "complete.obs")

total_plot = ggplot(combined_student_chapter_summary, aes(x = total_attempts, 
                         y = prop_correct, 
                         color = factor(chapter_num))) +
  geom_point() +
    labs(title = "",
       x = "Total Attempts",
       y = "Proportion items Correct",
       color = "Chapter") +
  theme_minimal() 

total_plot
cor(combined_student_chapter_summary$total_attempts, combined_student_chapter_summary$prop_correct, use = "complete.obs")

```

### 4) persistence  
```{r}
ggplot(items, aes(x = chapter_num, y = correctness_over_attempts, color = factor(chapter_num))) +
  geom_point(alpha = 0.3) +
  labs(title = "", 
      x = "Chapter", 
       y = "correctness over attempts") +
  theme_minimal() +
  theme(legend.position = "none")

# Box plot: distribution of avg. attempts per chapter
ggplot(items, aes(x = factor(chapter_num), y = correctness_over_attempts)) +
  geom_boxplot(fill = "lightblue", color = "darkblue", alpha = 0.1) +
  labs(title = "", 
       x = "Chapter", 
       y = "correctness over attempts") +
  theme_minimal()
```
```{r}
ggplot(items, aes(x = correctness_over_attempts)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "black") +  
  labs(title = "",
       x = "Correctness Over Attempts",
       y = "Count") +
  theme_minimal()
```
if never correct, how many attempts?
```{r}
ggplot(items |> filter(is_correct_ever != 1), aes(x = n_attempts)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +  
  labs(title = "",
       x = "num attempts when never correct",
       y = "Count") +
  theme_minimal()

favstats(~ n_attempts, data = items |> filter(is_correct_ever != 1))

ggplot(items |> filter(is_correct_ever == 1), aes(x = n_attempts)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +  
  labs(title = "",
       x = "num attempts when never correct",
       y = "Count") +
  theme_minimal()
favstats(~ n_attempts, data = items |> filter(is_correct_ever == 1))
```